Rising from the Ashes II: A Hidden Abundance Bimodality in Illustris TNG

### 2024-09-03 paper/ms.tex
Add follow up sentence
Update title
First pass changes

\shorttitle{Something Something}
\shortauthors{Beane et al.}

@@

\title{Rising from the Ashes II.: something something}

@@

\begin{abstract}
    The Milky Way is known to host at least two modes in the present day distribution of chemical abundances. The exact cause of this bimodality is disputed, but one class of explanations involves the merger between the Milky Way and a relatively massive sattelite (Gaia-Sausage-Enceladus) at $z\sim2$. However, reproducing this bimodality in simulations is not straightforward, with conflicting results on the prevalance, morphology, and mechanism behind multimodality. We present a case study of a galaxy in the Illustris TNG50 simulation which undergoes a starburst and brief quiescent phase. After a minor post-processing step which boosts the \alphaFe{} of old star particles, we demonstrate that this galaxy hosts a strongly bimodal distribution in the \alphaFe{}-\FeH{} plane. The high- and low-$\alpha$ sequences are neatly separated in time by the brief quiescent period. The quiescent period in this galaxy is not associated with a merger but some unspecified internal process which leads to AGN activity. We argue that the post-processing step can be understood as the TNG model underproducing SF in the densest regions at high-$z$.
  \end{abstract}

@@

\section{Introduction}\label{sec:intro}
For most elements, their stellar surface abundances retain the composition of their natal gas cloud. Therefore, the present-day distribution of stellar surface abundances encodes the chemical history of a galaxy. Two elements have received particular interest in the Milky Way: Fe and $\alpha$-elements (elements produced through the $\alpha$-process). Fe is produced in both Type Ia and Type II SNe whereas $\alpha$-elements are produced predominantly through Type II SNe. Because these SNe occur on different timescales (10s of Myr after star formation for Type II, as compared to 100s of Myr for Type Ia), the ratio between $\alpha$-elements and Fe is expected to generally decrease with time.

Because of their disparate formation channels, the two dimensional plane of \alphaFe{}-\FeH{} has received considerable interest. In the Milky Way, there is a well-established bimodality with separate high- and low-$\alpha$ sequences \citep{1996ASPC...92..307G,1998A&A...338..161F,2004AN....325....3F,2006MNRAS.367.1329R,2011A&A...535L..11A,2012A&A...545A..32A,2014A&A...562A..71B,2014ApJ...796...38N,2020MNRAS.493.2952H}. This bimodal distribution is shown in the upper left panel of Figure~\ref{fig:fig1}.

@@

In \citet{2024arXiv240707985B} we argued for an alternate formation scenario of the bimodality, for which we provide further support in this work. In this scenario, the bimodality is formed through a brief quiescent period in the Galaxy's history. During this period, which lasts $\sim300\,\Myr$, two things happen. First, as the chemical evolution of the galaxy proceeds fewer stars form. Second, because the SFR is lower, $\alpha$-element production drops and so the \alphaFe{} of the gas drops as well. The combination of these two effects is what leads to the valley in between the high- and low-$\alpha$ sequences.

In that work, we used idealized simulations of a galaxy merger that triggered a starburst which preceded the quiescent period. However, we argued that the merger aspect of that work was not necessary, but rather the quiescence. In this work, we study a subhalo from the Illustris TNG50 simulation which demonstrates this. This subhalo of interest (SoI) exhibits the sequence of events presented in \citet{2024arXiv240707985B} after a simple post-processing step which enhances the \alphaFe{} of old star particles. The SoI undergoes a brief quiescent period which neatly separates a high- and low-$\alpha$ sequence.

@@

\section{Methods}\label{sec:methods}
\subsection{IllustrisTNG Sample}\label{ssec:tng}
We have made use of the Illustris TNG50 simulation \citep{2019MNRAS.490.3196P, 2019MNRAS.490.3234N}, a cosmological simulation of a $\sim50\,\textrm{cMpc}$ box at high resolution ($m_{\textrm{baryon}}\sim8.5\times10^4\,\Msun$) It uses the gravito-magneto-hydrodynamics code \texttt{AREPO} \citep{2010MNRAS.401..791S, 2016MNRAS.455.1134P}, along with the TNG model \citep{2017MNRAS.465.3291W, 2018MNRAS.473.4077P}. This model includes several subgrid processes: a wind generation model, chemical enrichment from supernovae and asymptotic giant branch stars, and thermal and kinetic feedback from AGN.

Using the public catalog, we selected a sample of subhalos at $z=1.5$ (snapshot 40) according to the following criteria: (1) the subhalo is central (i.e., the most massive subhalo within its halo), and (2) the subhalo's stellar mass is between $10^{10}$ and $10^{10.5}\,\Msun/\textrm{h}$. There were a total of 168 subhalos that met both criteria. The chosen mass range is broadly consistent with the expected mass of the Milky Way at this redshift \citep{2013ApJ...771L..35V}. We chose to make our selection of galaxies at $z=1.5$ instead of at lower redshift because we wished to capture the \textit{formation} of any multimodal structure. We did not want contamination by mergers at lower redshift which we know contribute very little to the Milky Way's disk stars \citep[e.g.][]{2016ARA&A..54..529B}.

We examined the abundance distribution of this sample of subhalos by eye. Not many subhalos had multimodal structure, and any structure that was present was relatively weak compared to that observed in the Milky Way. We then applied a post-processing to the \MgFe{} of stars by adding to each star particle a value of $0.1\times\left(t_{1.5}-t_{\textrm{form}}\right)$, where $t_{1.5}$ is the age of the universe at $z=1.5$ ($\sim4.3\,\Gyr$) and $t_{\textrm{form}}$ is the formation time of each star particle. With this post-processing, we found that much more structure was generally present and we selected subhalo 172175 (at snapshot 40) for its particular resemblance to the Milky Way. We then studied the main descendant of this subhalo at $z=0$ (subhalo 392276 at snapshot 99). We refer to this subhalo as our subhalo of interest (SoI).

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{fig/392276.pdf}
  \caption{\textbf{When old stars are $\alpha$-enhanced, our subhalo of interest from TNG displays a prominent bimodality.} The upper left panel shows the distribution in the \MgFe{}-\FeH{} plane of the Milky Way, demonstrating a clear bimodality (data selection given in text). The lower left panel shows the 1D histograms of \MgFe{} at fixed \FeH{} values of $-0.5$, $-0.25$, $0$, and $0.25$ (blue, orange, green, and red, respectively). In the Milky Way, the bimodality is strongest at low metallicities while disappearing at high metallicities. The middle column shows the same plots but for our TNG subhalo of interest (392276) and with the fixed \FeH{} values $0.25\,\dex$ lower. No clear bimodality is detected at any metallicity. The right column shows the same subhalo but after increasing the \MgFe{} value of star particles formed before $z=1.5$ linearly with formation time (specifically by incrementing \MgFe{} by $0.1\times\left(t_{1.5}-t_{\textrm{form}}\right)$ if $t_{\textrm{form}} < t_{1.5}$, where $t_{1.5}$ is the age of the universe at $z=1.5$). A clear bimodality is shown in these panels which, unlike in the Milky Way, is present at all metallicities.}
  \label{fig:fig1}
\end{figure*}

The main result of our paper is given in Figure~\ref{fig:fig1}. Here, we compare the abundance plane in the Milky Way (left column) to that in our subhalo of interest (middle and right columns). The colored vertical lines show 1D conditional histograms at $\FeH=-0.75$, $-0.5$, $-0.25$, and $0\,\dex$ in the Milky Way, and at $0.25\,\dex$ higher in the simulations. The \MgFe{} values in the lower panels are given offsets in order to reduce overlap. The rationale for the higher plotted \FeH{} in the simulations comes just from the empirical location of the bimodalities. The Milky Way shows a clear bimodal population, with a high-$\alpha$ sequence most clearly distinct from the low-$\alpha$ sequence at low metallicity. The two sequences merge around solar metallicity.

Our SOI, on the other hand, does not show a clearly bimodal structure in the fiducial simulation (middle column). There is some structure in the $\FeH=-0.75$ bin, but at higher metallicity there is only a weak hint at a second mode in the $\FeH=-0.25$ bin.

The right panel of Figure~\ref{fig:fig1} shows the same distribution as in the middle panel, but with a post-processed declination in \MgFe{}. Star particles formed before $z=1.5$ are given an additive offset of $0.1\times\left(t_{1.5}-t_{\textrm{form}}\right)$, where $t_{1.5}$ is the age of the universe at $z=1.5$ and $t_{\textrm{form}}$ is the formation time of the star particle. Only stars formed before $z=1.5$ are shown. A multimodal structure emerges with three clear modes at $\MgFe\sim0.8$, $0.5$, and $0.2\,\dex$. The 1D histograms show that the modes are well-separated, and that the troughs between the modes nearly vanish.

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{fig/392276_alpha.pdf}
  \caption{\textbf{Bimodality in the abundance plane is linked to distinct epochs in simulation.} The upper panels show \MgFe{} as a function of age for our subhalo in TNG. The colors indicate stellar populations at fixed values of \FeH{}, which are the same as in Figure~\ref{fig:fig1}. A gap in the relation occurs at an age of approximately $10.75\,\Gyr$, which we indicate with a vertical dashed line. The effect of the $\alpha$-enhancement is clear, as it more clearly separates the stars that form before and after this gap in ages (star particles which formed before $z=1.5$ are $\alpha$-enhanced, which occurs at an age of $\sim9.5\,\Gyr$). The lower panels show on the left the Milky Way and on the center and right the data from TNG but with $10\%$ age errors and $0.01\,\dex$ errors in \MgFe{}. When the simulations are given these errors, we see that the before and after star particles smear such that the two populations significantly overlap in ages. This feature more closely resembles the Milky Way, which displays such populations where the bimodality is strongest -- $\FeH=-0.5$ (blue) and $-0.25$ (orange).}
  \label{fig:alpha}
\end{figure*}

The abundance distributions shown in Figure~\ref{fig:fig1} can be better understood by examining the evolution of \MgFe{} with time of the individual stars/star particles. In the upper panels of Figure~\ref{fig:alpha} we show the true distribution of \MgFe{} as a function of time for the fiducial SOI in the middle and for the post-processed, $\alpha$-enhanced subhalo to the right. We split  We use age instead of formation time in order to better facilitate comparisons to observations. There is a gap in the ages that occurs at an age of  $\sim10.75\,\Gyr$, which we mark with a vertical dashed line. Star particles older than this line have a much clearer gradient in time with \MgFe{} than stars that form after. In the $\FeH=-0.25$ bin, star particles which form directly after this line have a slightly depressed \MgFe{} than stars which form a short time later.

In the lower panels we plot to the left data from the Milky Way and to the middle and right from the fiducial and $\alpha$-enhanced SOI, respectively. For the simulations, we add $10\%$ age and $0.01\,\dex$ \MgFe{} errors. These values are roughly in line with the observational errors from the APOKASC2 and APOGEE datasets that the lower left panel is made from (see Appendix~\ref{app:errors} for plots of the observational errors). The error in \MgFe{} is not signficant, but the age error of $1\,\Gyr$ at $10\,\Gyr$ smears out the distribution before and after the dashed line. The $\alpha$-enhanced SOI still shows two separate populations in this plot, but they now significantly overlap in age.

The Milky Way distribution (lower left panel) bears some resemblance to the $\alpha$-enhanced SOI. In particular, the $\FeH=-0.5$ and $-0.25$ bins (blue and orange, respectively) show what appears to be two populations overlapping in age but nonetheless separated in \MgFe{}.

@@

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{fig/392276_SFH_AGN_M200.pdf}
  \caption{\textbf{The evolutionary history of our subhalo of interest.} The upper panel shows the SFH of our subhalo. This SFH is generated using the formation times of all star particles in the subhalo at $z=0$. In this and all subsequent panels we show a vertical dashed line at the same position as in Figure~\ref{fig:alpha}, which delineates the separation between the high- and low-$\alpha$ sequences. The gap in ages is naturally associated with a gap in the SFH located at the dashed line. The middle panel shows the BH accretion rate as a fraction of the maximum (Eddington) accretion rate across time. At early times the accretion rate is high, but the gap in ages is also coincident with a local spike in the accretion rate. In the lower panel we show the mass assembly of the halo, given as the mass within the radius that encloses $200\times$ the mean density of the universe. In this plot mergers are shown as sudden increases in the halo mass. Mergers at very early times when the halo is signifcantly less massive than its $z=0$ mass are not shown, but one can see that no clear merger is associated with the gap in stellar ages. Two mergers occur later on at $t\sim9$ and $\sim12\,\Gyr$.}
  \label{fig:history}
\end{figure}

In an effort towards understanding the events in our SOI's history that led to the behavior around the dashed line in Figure~\ref{fig:alpha}, we examine the evolution of some of its key quantities. In the upper panel of Figure~\ref{fig:history}, we show the SFH of our galaxy with the same dashed line as in Figure~\ref{fig:alpha} (at an age of $10.75\,\Gyr$ in that plot is at a time of $\sim3\,\Gyr$ here). This SFH is computed from the formation times of all star particles in the subhalo at $z=0$. This allows us to compute the SFH at fine time resolution, but also means contamination by accreted stellar debris. We see two peaks in the SFH at $t\sim2.5$ and $\sim4.5\,\Gyr$ of an amplitude of about $70$ and $60\,\Msunyr$. The dashed line corresponds to a relative drop in the SFR, down to about $10\,\Msunyr$ before quickly recovering.

The middle panel shows the accretion rate of the central black hole as a fraction of the maximum (Eddington) accretion rate. Early in its history ($t<2\,\Gyr$), the subhalo experiences high accretion rates. The accretion rate then steadily declines until $t\sim5\,\Gyr$ at which point it switches to the low accretion/radio mode. However, at the dashed line there is a localized peak in the accretion rate, maxing out at about $30\%$ Eddington.

@@

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{fig/mgfe_vice.pdf}
  \caption{\textbf{A higher star formation efficiency leads to a steeper decline in \MgFe{}.} In both panels, the lines show the time evolution of \MgFe{} in a simple one zone chemical evolution model, described in the text. The upper panel shows the evolution over $2\,\Gyr$ of \MgFe{} while the lower panel shows the negative of its time derivative. Decreasing the star formation timescale $\tau_{\star}=M_{\textrm{gas}/\textrm{SFR}}$ leads to a more rapid decline in \MgFe{}. At its steepest decline ($t\sim0.5\,\Gyr$), an order of magnitude decrease in $\tau_{\star}$ leads to a slope nearly a factor of $2$ larger. At later times ($t>1\,\Gyr$), the models with smaller $\tau_{\star}$ reach their steady-state \MgFe{} value more quickly.}
  \label{fig:vice}
\end{figure}

@@

We show in the upper panel the evolution of \MgFe{} of three one zone chemical evolution models which vary the star formation timescale. This timescale, $\tau_{\star}$, is the inverse of the star formation efficiency. A shorter star formation timescale leads to a more rapid reduction in \MgFe{}. In the shortest timescale model, $\tau_{\star}=0.5\,\Gyr$, \MgFe{} drops from $\sim0.45$ to $0.08\,\dex$ in $2\,\Gyr$. In comparison, the longest timescale model, $\tau_{\star}=5\,\Gyr$ only drops to $\sim0.2\dex$ in the same time.

The lower panel shows the negative of the time derivative of \MgFe{}. This panel shows the same behavior, with the slope of the $\tau_{\star}=0.5\,\Gyr$ model peaking at $-0.5\,\dex/\Gyr$ compared to the $5\,\Gyr$ model which peaks at $-0.25\,\dex/\Gyr$. After $1\,\Gyr$, the trend starts to reverse, with the longer star formation timescale models declining more rapidly, though at a much reduced rate ($\sim-0.1\,\dex/\Gyr$) than at the peak.

\section{Discussion}\label{sec:disc}
In Figure~\ref{fig:fig1}, we compared the abundance plane between the Milky Way and our SOI before and after the $\alpha$-enhancement. It is visibly obvious that the TNG SOI is unimodal before the $\alpha$-enhancement and bimodal afterwards (ignoring the minor mode at high-\MgFe{}). Here, we briefly discuss two main points: (1) assuming the $\alpha$-enhancement is justified, what leads to the bimodality in the SOI?, and (2) what justifies the $\alpha$-enhancement? We then briefly extend our comparison to the Milky Way data.

\subsection{Cause of Bimodality}\label{ssec:bim_cause}
An extensive analysis of our SOI is beyond the scope of this project, but here we argue that the evidence is consistent with the scenario presented in \citet{2024arXiv240707985B}. They presented an idealized simulation resembling the merger between the Milky Way and GSE. They varied the orbital parameters slightly in a grid of 27 simulations and found that simulations which had a brief quiescent period as a result of the merger led to a bimodal abundance distribution. They argued that AGN activity was responsible for this brief quenching.

@@

The gap in each \FeH{} bin of Figure~\ref{fig:alpha} is contemperaneous with a minimum in the global SFR. We can see in the upper panel of Figure~\ref{fig:history} that this dashed line lies almost exactly at the point of a local minimum in the SFH. This minimum, which is $\sim10\,\Msunyr$, is 6--7$\times$ smaller than the maxima before and after it.\footnote{A reasonable argument can be made that $10\,\Msunyr$ is hardly quiescent. However, for our purposes, it only matters that the SFR at the minimum is small relative to the SFR close in time before and after. And besides, this SFH is computed using the formation times of all star particles at $z=0$, which is likely contaminated by later mergers.}

@@

The reduction in the SFR has two effects. First, a lower production of Mg is the natural result of a lack of Type~II SNe. Second, the lack of SF means that stars at \MgFe{} intermediate between the high- and low-$\alpha$ sequences never form. This is the same argument that was first presented in \citet{2024arXiv240707985B}. When observational errors are added to the subhalo's abundance vs age distribution, we show that it is broadly consistent with the Milky Way's distribution (Section~\ref{ssec:compare_obs}).

@@

% \input{src/app.tex}

\end{document}

>>>>

\shorttitle{Abundance Bimodality in TNG}
\shortauthors{Beane et al.}

@@

\title{Rising from the Ashes II: A Hidden Abundance Bimodality in Illustris TNG}

@@

\begin{abstract}
    The Milky Way is known to host at least two modes in the present day distribution of Fe and $\alpha$-elements. The exact cause of this bimodality is disputed, but one class of explanations involves the merger between the Milky Way and a relatively massive sattelite (Gaia-Sausage-Enceladus) at $z\sim2$. However, reproducing this bimodality in simulations is not straightforward, with conflicting results on the prevalance, morphology, and mechanism behind multimodality. We present a case study of a galaxy in the Illustris TNG50 simulation which undergoes a sequence of starburst, brief quiescence, and then rejuvenation. After a minor post-processing step which boosts the \alphaFe{} of old star particles, we demonstrate that this galaxy hosts a strongly bimodal distribution in the \alphaFe{}-\FeH{} plane. The high- and low-$\alpha$ sequences are neatly separated in time by the brief quiescent period. The quiescent period in this galaxy is not associated with a merger but some unspecified internal process which leads to AGN activity. We argue that the post-processing step can be understood as the TNG model underproducing star formation in the densest regions at high-$z$.
  \end{abstract}

@@

\section{Introduction}\label{sec:intro}
The stellar surface abundances of most elements retain the composition of their natal gas cloud. Therefore, the present-day distribution of stellar surface abundances encodes the chemical history of a galaxy's gas phase. Two elements have received particular interest in the Milky Way: Fe and $\alpha$-elements (elements produced through the $\alpha$-process, such as O and Mg). Fe is produced in both Type Ia and Type II SNe whereas $\alpha$-elements are produced predominantly through Type II SNe. Because these SNe occur on different timescales (10s of Myr after star formation for Type II, as compared to 100s of Myr to Gyrs for Type Ia), the ratio between $\alpha$-elements and Fe is expected to generally decrease with time.

Because of their separate formation channels, the two dimensional plane of \alphaFe{}-\FeH{} has received considerable interest. In the Milky Way, there is a well-established bimodality with separate high- and low-$\alpha$ sequences \citep{1996ASPC...92..307G,1998A&A...338..161F,2004AN....325....3F,2006MNRAS.367.1329R,2011A&A...535L..11A,2012A&A...545A..32A,2014A&A...562A..71B,2014ApJ...796...38N,2020MNRAS.493.2952H}. This bimodal distribution is shown in the upper left panel of Figure~\ref{fig:fig1}.

@@

In \citet{2024arXiv240707985B} we argued for an alternate formation scenario of the bimodality, for which we provide further support in this work. In this scenario, the bimodality is formed through a brief quiescent period in the Galaxy's history. During this period, which lasted $\sim300\,\Myr$ in that setup, two things happen. First, as the chemical evolution of the galaxy proceeds, fewer stars form. Second, because the SFR is lower, $\alpha$-element production drops and so the \alphaFe{} of the gas drops as well. The combination of these two effects is what leads to the valley in between the high- and low-$\alpha$ sequences.

In that work, we used idealized simulations of a galaxy merger that triggered a starburst which preceded the quiescent period. However, we argued that the merger aspect of that work was not necessary, but rather the quiescence. In this work, we study a subhalo from the Illustris TNG50 simulation which demonstrates this. This subhalo of interest (SoI) exhibits the sequence of events presented in \citet{2024arXiv240707985B} after a simple post-processing step which enhances the \alphaFe{} of old star particles. The SoI undergoes a brief quiescent period which neatly separates a high- and low-$\alpha$ sequence. Therefore, this work serves as a verification that the scenario in \citet{2024arXiv240707985B} is possible in cosmological simulations and can occur in non-merger scenarios.

@@

\section{Methods}\label{sec:methods}
\subsection{IllustrisTNG Sample}\label{ssec:tng}
We have made use of the Illustris TNG50 simulation \citep{2019MNRAS.490.3196P, 2019MNRAS.490.3234N}, a cosmological simulation of a $\sim50\,\textrm{cMpc}$ box at high resolution ($m_{\textrm{baryon}}\sim8.5\times10^4\,\Msun$). It uses the gravito-magneto-hydrodynamics code \texttt{AREPO} \citep{2010MNRAS.401..791S, 2016MNRAS.455.1134P}, along with the TNG model \citep{2013MNRAS.436.3031V, 2017MNRAS.465.3291W, 2018MNRAS.473.4077P}. This model includes several subgrid processes: a wind generation model, chemical enrichment from supernovae and asymptotic giant branch stars, and thermal and kinetic feedback from AGN.

Using the public catalog, we selected a sample of subhalos at $z=1.5$ (snapshot 40) according to the following criteria: (1) the subhalo is central (i.e., the most massive subhalo within its halo), and (2) the subhalo's stellar mass is between $10^{10}$ and $10^{10.5}\,\Msun/\textrm{h}$. There were a total of 168 subhalos that met both criteria. The chosen mass range is broadly consistent with the expected mass of the Milky Way at this redshift \citep{2013ApJ...771L..35V}. We chose to make our selection of galaxies at $z=1.5$ instead of at lower redshift because we wished to capture the \textit{formation} of any multimodal structure. We did not want contamination by mergers at lower redshift which we know contribute very little to the Milky Way's disk stars \citep[e.g.,][]{2016ARA&A..54..529B}.

We examined the abundance distribution in the \MgFe{}-\FeH{} plane of this sample of subhalos by eye. Not many subhalos had multimodal structure, and any structure that was present was relatively weak compared to that observed in the Milky Way. We then applied a post-processing to the \MgFe{} of stars by adding to each star particle a value of $0.1\times\left(t_{1.5}-t_{\textrm{form}}\right)$, where $t_{1.5}$ is the age of the universe at $z=1.5$ ($\sim4.3\,\Gyr$) and $t_{\textrm{form}}$ is the formation time of each star particle. With this post-processing, we found that much more structure was generally present. We selected subhalo 172175 (at snapshot 40) for its particular resemblance to the Milky Way. We then studied the main descendant of this subhalo at $z=0$ (subhalo 392276 at snapshot 99). We refer to this subhalo as our subhalo of interest (SoI).

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{fig/392276.pdf}
  \caption{\textbf{When old stars are $\alpha$-enhanced, our subhalo of interest from TNG displays a prominent bimodality.} The upper left panel shows the distribution in the \MgFe{}-\FeH{} plane of the Milky Way, demonstrating a clear bimodality (data selection given in text). The lower left panel shows the 1D histograms of \MgFe{} at fixed \FeH{} values of $-0.5$, $-0.25$, $0$, and $0.25$ (blue, orange, green, and red, respectively). In the Milky Way, the bimodality is strongest at low metallicities while disappearing at high metallicities. The middle column shows the same plots but for our TNG subhalo of interest (392276) and with the fixed \FeH{} values $0.25\,\dex$ lower. Only faint structure is seen in the lowest bin (blue, $-0.75\,\dex$). The right column shows the same subhalo but after increasing the \MgFe{} value of star particles formed before $z=1.5$ linearly with formation time (specifically by incrementing \MgFe{} by $0.1\times\left(t_{1.5}-t_{\textrm{form}}\right)$ if $t_{\textrm{form}} < t_{1.5}$, where $t_{1.5}$ is the age of the universe at $z=1.5$). A clear bimodality is shown in these panels which, unlike in the Milky Way, is present at all metallicities.}
  \label{fig:fig1}
\end{figure*}

The main result of our paper is given in Figure~\ref{fig:fig1}. Here, we compare the abundance plane in the Milky Way (left column) to that in our subhalo of interest (middle and right columns). The colored vertical lines show 1D conditional histograms at $\FeH=-0.75$, $-0.5$, $-0.25$, and $0\,\dex$ in the Milky Way, and at bins $0.25\,\dex$ higher in the simulations. The \MgFe{} values in the lower panels are given offsets in order to reduce overlap. The rationale for the higher plotted \FeH{} in the simulations comes just from the empirical location of the bimodalities. The Milky Way shows a clear bimodal population, with a high-$\alpha$ sequence most clearly distinct from the low-$\alpha$ sequence at low metallicity. The two sequences merge around solar metallicity.

Our SoI, on the other hand, does not show a clearly bimodal structure in the fiducial simulation (middle column). There is some structure in the $\FeH=-0.75$ bin. The right panel of Figure~\ref{fig:fig1} shows the same distribution as in the middle panel, but with a post-processed declination in \MgFe{} described in Section~\ref{ssec:tng}. Star particles formed before $z=1.5$ are given an additive offset of $0.1\times\left(t_{1.5}-t_{\textrm{form}}\right)$, where $t_{1.5}$ is the age of the universe at $z=1.5$ and $t_{\textrm{form}}$ is the formation time of the star particle. A multimodal structure emerges with three clear modes at $\MgFe\sim0.8$, $0.5$, and $0.2\,\dex$. The 1D histograms show that the modes are well-separated, and that the troughs between the modes nearly vanish.

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{fig/392276_alpha.pdf}
  \caption{\textbf{Bimodality in the abundance plane is linked to distinct epochs in simulation.} The upper panels show \MgFe{} as a function of age for our subhalo in TNG. The colors indicate stellar populations at fixed values of \FeH{}, which are the same as in Figure~\ref{fig:fig1}. A gap in the relation occurs at an age of approximately $10.75\,\Gyr$, which we indicate with a vertical dashed line. The effect of the $\alpha$-enhancement is clear, as it separates the stars that form before and after this gap in ages (star particles which formed before $z=1.5$ are $\alpha$-enhanced, which occurs at an age of $\sim9.5\,\Gyr$). The lower panels show on the left the Milky Way and on the center and right the data from TNG but with $15\%$ age errors and $0.01\,\dex$ errors in \MgFe{}. When the simulations are given these errors, we see that the before and after star particles smear such that the two populations significantly overlap in ages. In the $\alpha$-enhanced SoI, two populations emerge in each bin which overlapped in the fiducial distribution. This feature more closely resembles the Milky Way, which displays such populations where the bimodality is strongest -- $\FeH=-0.5$ (blue) and $-0.25$ (orange).}
  \label{fig:alpha}
\end{figure*}

The abundance distributions shown in Figure~\ref{fig:fig1} can be better understood by examining the evolution of \MgFe{} with time of the individual stars/star particles. In the upper panels of Figure~\ref{fig:alpha} we show the true distribution of \MgFe{} as a function of time for the fiducial SoI in the middle and for the post-processed, $\alpha$-enhanced subhalo to the right. We use age instead of formation time in order to better facilitate comparisons to observations. There is a gap in the ages that occurs at an age of $10.75\,\Gyr$, which we mark with a vertical dashed line. Star particles older than this line have a much clearer gradient in time with \MgFe{} than stars that form after. In the $\FeH=-0.25$ bin, star particles which form directly after this line have a slightly depressed \MgFe{} than stars which form a short time later.

In the lower panels we plot to the left data from the Milky Way and to the middle and right from the fiducial and $\alpha$-enhanced SoI, respectively. For the simulations, we add $15\%$ age and $0.01\,\dex$ \MgFe{} Gaussian errors. These values are roughly in line with the observational errors from the APOKASC2 and APOGEE datasets that the lower left panel is made from (see Appendix~\ref{app:errors} for plots of the observational errors). The error in \MgFe{} is not signficant, but the age error, which is $1\,\Gyr$ at $10\,\Gyr$, smears out the distribution before and after the dashed line. The $\alpha$-enhanced SoI still shows two separate populations in this plot, but they now significantly overlap in age.

The Milky Way distribution (lower left panel) bears some resemblance to the $\alpha$-enhanced SoI. In particular, the $\FeH=-0.5$ and $-0.25$ bins (blue and orange, respectively) show what appears to be two populations overlapping in age but nonetheless separated in \MgFe{}. These are the bins at which the bimodality is strongest (upper left panel of Figure~\ref{fig:fig1}).

@@

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{fig/392276_SFH_AGN_M200.pdf}
  \caption{\textbf{The evolutionary history of our subhalo of interest.} The upper panel shows the SFH of our subhalo. This SFH is generated using the formation times of all star particles in the subhalo at $z=0$. In this and all subsequent panels we show a vertical dashed line at the same position as in Figure~\ref{fig:alpha} ($\sim3\,\Gyr$), which delineates the separation between the high- and low-$\alpha$ sequences. The gap in ages is naturally associated with a gap in the SFH located at the dashed line. The middle panel shows the BH accretion rate as a fraction of the maximum (Eddington) accretion rate across time. The accretion rate is high early on, but then drops. The gap in ages (dashed line) is coincident with a local peak in the BH accretion rate. In the lower panel we show the mass assembly of the halo, given as the mass within the radius that encloses $200\times$ the mean density of the universe. In this plot mergers are shown as sudden increases in $M_{200}$. Mergers at very early times when the halo is signifcantly less massive than its $z=0$ mass are not shown, but one can see that no clear merger is associated with the gap in stellar ages. Two mergers occur later on at $t\sim10$ and $\sim12\,\Gyr$.}
  \label{fig:history}
\end{figure}

In an effort towards understanding the events in our SoI's history that led to the behavior around the dashed line in Figure~\ref{fig:alpha}, we examine the evolution of some of its key quantities. In the upper panel of Figure~\ref{fig:history}, we show the SFH of our galaxy with the same dashed line as in Figure~\ref{fig:alpha} (at an age of $10.75\,\Gyr$ in that plot is at a time of $\sim3\,\Gyr$ here). This SFH is computed from the formation times of all star particles in the subhalo at $z=0$. This allows us to compute the SFH at fine time resolution. We see two peaks in the SFH at $t\sim2.5$ and $\sim4.5\,\Gyr$ of an amplitude of about $70$ and $60\,\Msunyr$. The dashed line corresponds to a relative drop in the SFR, down to about $10\,\Msunyr$ before quickly recovering.

The middle panel shows the accretion rate of the central black hole as a fraction of the maximum (Eddington) accretion rate. Early in its history ($t<2\,\Gyr$), the subhalo experiences high accretion rates. The accretion rate then steadily declines until $t\sim5\,\Gyr$ at which point it switches to the low accretion radio mode. However, at the dashed line ($\sim3\,\Gyr$) there is a localized peak in the accretion rate, maxing out at about $30\%$ Eddington.

@@

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{fig/mgfe_vice.pdf}
  \caption{\textbf{A higher star formation efficiency leads to a steeper decline in \MgFe{}.} In both panels, the lines show the time evolution of \MgFe{} in a simple one zone chemical evolution model, described in Section~\ref{ssec:onezone_met}. The upper panel shows the evolution of \MgFe{} over $2\,\Gyr$ while the lower panel shows the negative of its time derivative. Decreasing the star formation timescale $\tau_{\star}=M_{\textrm{gas}/\textrm{SFR}}$ leads to a more rapid decline in \MgFe{}. At its steepest decline ($t\sim0.5\,\Gyr$), an order of magnitude decrease in $\tau_{\star}$ leads to a slope nearly a factor of $2$ larger. At later times ($t>1\,\Gyr$), the models with smaller $\tau_{\star}$ reach their steady-state \MgFe{} value more quickly.}
  \label{fig:vice}
\end{figure}

@@

We show in the upper panel the evolution of \MgFe{} of three one zone chemical evolution models which vary the star formation (SF) timescale. This timescale, $\tau_{\star}$, is the inverse of the SF efficiency. A shorter SF timescale leads to a more rapid reduction in \MgFe{}. In the shortest timescale model, $\tau_{\star}=0.5\,\Gyr$, \MgFe{} drops from $\sim0.45$ to $0.08\,\dex$ in $2\,\Gyr$. In comparison, the longest timescale model, $\tau_{\star}=5\,\Gyr$ only drops to $\sim0.2\,\dex$ in the same time.

The lower panel shows the negative of the time derivative of \MgFe{}. This panel shows the same behavior, with the slope of the $\tau_{\star}=0.5\,\Gyr$ model peaking at $-0.5\,\dex/\Gyr$ compared to the $5\,\Gyr$ model which peaks at $-0.25\,\dex/\Gyr$. After $1\,\Gyr$, the trend starts to reverse, with the longer SF timescale models declining more rapidly, though at a much reduced rate ($\sim-0.1\,\dex/\Gyr$) than at the peak.

\section{Discussion}\label{sec:disc}
In Figure~\ref{fig:fig1}, we compared the abundance plane between the Milky Way and our SoI before and after the $\alpha$-enhancement. It is visibly obvious that the TNG SoI is unimodal before the $\alpha$-enhancement and bimodal afterwards (ignoring the minor mode at high-\MgFe{}). Here, we briefly discuss two main points: (1) assuming the $\alpha$-enhancement is justified, what leads to the bimodality in the SoI?, and (2) what justifies the $\alpha$-enhancement? We then briefly extend our comparison to the Milky Way data.

\subsection{Cause of Bimodality}\label{ssec:bim_cause}
An extensive analysis of our SoI is beyond the scope of this project, but here we argue that the evidence is consistent with the scenario presented in \citet{2024arXiv240707985B}. They presented an idealized simulation resembling the merger between the Milky Way and GSE. They varied the orbital parameters slightly in a grid of 27 simulations and found that simulations which had a brief quiescent period as a result of the merger led to a bimodal abundance distribution. They argued that AGN activity was responsible for this brief quenching.

@@

The age gap in each \FeH{} bin of Figure~\ref{fig:alpha} is contemperaneous with a minimum in the global SFR. We can see in the upper panel of Figure~\ref{fig:history} that this dashed line lies almost exactly at the point of a local minimum in the SFH. This minimum, which is $\sim10\,\Msunyr$, is 6--7$\times$ smaller than the maxima before and after it.\footnote{A reasonable argument can be made that $10\,\Msunyr$ is hardly quiescent. However, for our purposes, it only matters that the SFR at the minimum is small relative to the SFR close in time before and after.}

@@

The reduction in the SFR has two effects. First, a lower production of Mg is the natural result of a lack of Type~II SNe. Second, the lack of SF means that stars at \MgFe{} intermediate between the high- and low-$\alpha$ sequences never form. This is the same argument that was first presented in \citet{2024arXiv240707985B}. When observational errors are added to the subhalo's abundance vs age distribution, we show that it is broadly consistent with the Milky Way's abundance-age distribution (Section~\ref{ssec:compare_obs}).

@@

% \input{src/app.tex}
\end{document}
